The hedging hyperprior for the normal distribution: known mean, unknown precision

### 2023-03-18 text/23/normal-hedging-hyperprior-part-1.Rmd
Added gamma distribution to normal-heding-prior-part-2
Converted formulas in normal-hedging-hyperprior to use theta

### A simple Baysian analysis with an informative prior.

@@

@@

$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$

@@

$\tau=\frac{1}{\sigma*2}$

@@

Assume that $\tau$ is known and equal to $\tau_F$. A more general case, with $\tau$ unknown is described on a separate web page. For simplicity, we will drop any multiplicative constants not involving X or $\mu$

$f(X; \mu) \propto e^{-\frac{1}{2}\tau_F(X-\mu)^2}$

@@

$g(\mu; \mu_0, \tau_0) \propto e^{- \frac{1}{2}\tau_0(\mu-\mu_0)^2}$

where the prior mean is $\mu_0$ and the prior precision is $\tau_0$.

For an informative prior distribution, $\tau_0$ is usually much smaller than $\tau_F$. The ratio $n_0=\tau_F/\tau_0$ is the effective prior sample size and is equivalent to adding $n_0$ pseudo data values of $\mu_0$ to the observed data.

@@

$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F\sum(X_i-\mu)^2}$

@@

$L(\mathbf{X} | \mu) \propto e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\mu)^2}$

@@

$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2} e^{- \frac{n}{2} \tau_F (\bar{X}-\mu)^2}$

$L(\mathbf{X} | \mu) \propto e^{-\frac{n-1}{2}\tau_F S^2} e^{-\frac{n}{2} \tau_F (\mu-\bar{X})^2}$

@@

$h(\mu | \mathbf{X}) \propto e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2} e^{- \frac{n-1}{2} \tau_F S^2} e^{-\frac{n}{2} \tau_F(\mu-\bar{X})^2)}$

Drop the middle exponential term, which is constant with respect to $\mu$ and use high school algebra ([completing the square][wik1]) to get

$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$

@@

$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0}$

@@

$\frac{n \tau_F}{n\tau_F+\tau_0}\bar{X} + \frac{\tau_0}{n\tau_F+\tau_0}\mu_0$

@@

mu0 <- 4
tau0 <- 8

@@

mu1 <- (tau0*mu0+n*tauF*xbar)/(tau0+n*tauF)
tau1 <- tau0+n*tauF

@@

f <- function(x, mu, tau) {
  sqrt(tau/2*pi)*exp(-0.5*tau*(x-mu)^2)
}

@@

prior05 <- qnorm(0.025, mu0, 1/sqrt(tau0)) 
prior95 <- qnorm(0.975, mu0, 1/sqrt(tau0))

post05 <- qnorm(0.025, mu1, 1/sqrt(tau1)) 
post95 <- qnorm(0.975, mu1, 1/sqrt(tau1))

@@

Let's illustrate this with an informative prior with $\mu_0$ = `r mu0` and $\tau_0$ = `r tau0`.

```{r}
g_grid <- g(x_grid, mu0, tau0)
outside <- x_grid < prior05 | x_grid > prior95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=g_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

@@

```{r}
h_grid <- g(x_grid, mu1, tau1)
outside <- x_grid < post05 | x_grid > post95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=h_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x_grid, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r mu0` and it excludes `r xbar`, the mean of the data.

@@

Define your prior distribution as a normal distributions with prior mean $\mu0$ and prior precision

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

gstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF)+tauF)}
density <- outer(mu_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

This is the joint prior distribution of $\mu_0$ and $\lambda$.

It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\mu$ is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r}
density <- outer(mu_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
likelihood <- outer(mu_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
density <- outer(mu_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  mu_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(50, 75, 90)/100,
  xlab="mu",
  ylab="lambda")
```

You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r mu0`) is largely ignored in favor of the mean of the data (`r xbar`).

>>>>

### A simple Bayesian analysis with an informative prior.

@@

$f(X; \theta, \sigma) = \frac{\tau}{\sigma \sqrt{2 \pi}}e^{-(X-\theta)^2 / 2\sigma^2}$

@@

$f(X; \theta, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-\frac{1}{2}\tau (X-\theta)^2}$

@@

$\tau=\frac{1}{\sigma^2}$

@@

Assume that $\tau$ is known and equal to $\tau_F$. A more general case, with $\tau$ unknown is described on a separate web page. For simplicity, we will drop any multiplicative constants not involving X or $\theta$

$f(X; \theta) \propto e^{-\frac{1}{2}\tau_F(X-\theta)^2}$

@@

$g(\theta; \theta_0, \tau_0) \propto e^{- \frac{1}{2}\tau_0(\theta-\theta_0)^2}$

where the prior mean is $\theta_0$ and the prior precision is $\tau_0$.

For an informative prior distribution, $\tau_0$ is usually much smaller than $\tau_F$. The ratio $n_0=\tau_F/\tau_0$ is the effective prior sample size and is equivalent to adding $n_0$ pseudo data values of $\theta_0$ to the observed data.

@@

$L(\mathbf{X} | \theta) \propto e^{- \frac{1}{2} \tau_F\sum(X_i-\theta)^2}$

@@

$L(\mathbf{X} | \theta) \propto e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\theta)^2}$

@@

$L(\mathbf{X} | \theta) \propto e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2} e^{- \frac{n}{2} \tau_F (\bar{X}-\theta)^2}$

$L(\mathbf{X} | \theta) \propto e^{-\frac{n-1}{2}\tau_F S^2} e^{-\frac{n}{2} \tau_F (\theta-\bar{X})^2}$

@@

$h(\theta | \mathbf{X}) \propto e^{- \frac{1}{2} \tau_0 (\theta-\theta_0)^2} e^{- \frac{n-1}{2} \tau_F S^2} e^{-\frac{n}{2} \tau_F(\theta-\bar{X})^2)}$

Drop the middle exponential term, which is constant with respect to $\theta$ and use high school algebra ([completing the square][wik1]) to get

$h(\theta | \mathbf{X}) \propto e^{-(n\tau_F+\tau_0)(\theta-\frac{n \tau_F \bar{X} + \tau_0 \theta_0}{n\tau_F+\tau_0})^2}$

@@

$\frac{n \tau_F \bar{X} + \tau_0 \theta_0}{n\tau_F+\tau_0}$

@@

$\frac{n \tau_F}{n\tau_F+\tau_0}\bar{X} + \frac{\tau_0}{n\tau_F+\tau_0}\theta_0$

@@

theta0 <- 4
tau0 <- 8

@@

theta1 <- (tau0*theta0+n*tauF*xbar)/(tau0+n*tauF)
tau1 <- tau0+n*tauF

@@

f <- function(x, theta, tau) {
  sqrt(tau/2*pi)*exp(-0.5*tau*(x-theta)^2)
}

@@

prior05 <- qnorm(0.025, theta0, 1/sqrt(tau0)) 
prior95 <- qnorm(0.975, theta0, 1/sqrt(tau0))

post05 <- qnorm(0.025, theta1, 1/sqrt(tau1)) 
post95 <- qnorm(0.975, theta1, 1/sqrt(tau1))

@@

Let's illustrate this with an informative prior with $\theta_0$ = `r theta0` and $\tau_0$ = `r tau0`.

```{r}
g_grid <- g(x_grid, theta0, tau0)
outside <- x_grid < prior05 | x_grid > prior95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=g_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

@@

```{r}
h_grid <- g(x_grid, theta1, tau1)
outside <- x_grid < post05 | x_grid > post95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=h_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x_grid, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r theta0` and it excludes `r xbar`, the mean of the data.

@@

Define your prior distribution as a normal distributions with prior mean $\theta0$ and prior precision

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

gstar <- function(theta, lambda) {g(theta, theta0, lambda*(tau0-tauF)+tauF)}
density <- outer(theta_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

This is the joint prior distribution of $\theta_0$ and $\lambda$.

It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\theta$ is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r}
density <- outer(theta_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

fstar <- function(theta, lambda) {f(xbar, theta, tau0)^n}
likelihood <- outer(theta_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

hstar <- function(theta, lambda) {g(theta, theta0, lambda*(tau0-tauF) + tauF)*f(xbar, theta, tauF)^n}
density <- outer(theta_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  theta_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(50, 75, 90)/100,
  xlab="theta",
  ylab="lambda")
```

You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r theta0`) is largely ignored in favor of the mean of the data (`r xbar`).

### 2023-03-18 text/23/normal-hedging-hyperprior-part-2.Rmd
Converted formulas in normal-hedging-hyperprior to use theta

Consider a Bayesian analysis of an outcome variable that has a normal distribution. The density of the normal distribution is

$f(X; \mu, \sigma) = \frac{\tau}{\sigma sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$

For this derivation, it is easier to write the density in terms of

$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-\frac{1}{2}\tau (X-\mu)^2}$

@@

$\tau=\frac{1}{\sigma*2}$

The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance,

Assume that $\mu$ is known and equal to $\mu_F$. An earlier case with $\tau$ known and $\mu$ unknown appears on a different web page. A more general case, with both $\mu$ and $\tau$ unknown is also described on a separate web page. For simplicity, we will drop any multiplicative constants not involving X or $\tau$

$f(X; \tau) \propto \sqrt{\tau}e^{-\frac{1}{2}\tau(X-\mu_F)^2}$

@@

$g(\tau; \alpha_0, \beta_0) \propto \frac{\beta_0^\alpha_0}{\Gamma(\alpha)} \tau^{\alpha-1} e^{-\beta \tau}$

$\beta_0 = 1 / \tau_0$

For an informative prior distribution, $\alpha$ is large. This prior is is equivalent to adding $n_0$ pseudo data values of $\tau_0$ to the observed data.

@@

$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F\sum(X_i-\mu)^2}$

Rewrite this by incorporating $\bar{X}$ into the squared term

$L(\mathbf{X} | \mu) \propto e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\mu)^2}$

When you square the term inside the sum, the cross-product cancels out, leaving you with

$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2} e^{- \frac{n}{2} \tau_F (\bar{X}-\mu)^2}$

$L(\mathbf{X} | \mu) \propto e^{-\frac{n-1}{2}\tau_F S^2} e^{-\frac{n}{2} \tau_F (\mu-\bar{X})^2}$

@@

$h(\mu | \mathbf{X}) \propto e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2} e^{- \frac{n-1}{2} \tau_F S^2} e^{-\frac{n}{2} \tau_F(\mu-\bar{X})^2)}$

Drop the middle exponential term, which is constant with respect to $\mu$ and use high school algebra ([completing the square][wik1]) to get

$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$

which is a normal distribution. The posterior mean

$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0}$

and the posterior precision is

$n \tau_F + \tau_0$

The posterior mean can be written as a weighted average of the prior mean and the mean of the observed data.

$\frac{n \tau_F}{n\tau_F+\tau_0}\bar{X} + \frac{\tau_0}{n\tau_F+\tau_0}\mu_0$

@@

mu0 <- 4
tau0 <- 8

tauF <- 0.6
n <- 10
xbar <- 2

mu1 <- (tau0*mu0+n*tauF*xbar)/(tau0+n*tauF)
tau1 <- tau0+n*tauF

@@

f <- function(x, mu, tau) {
  sqrt(tau/2*pi)*exp(-0.5*tau*(x-mu)^2)
}

g <- f

prior05 <- qnorm(0.025, mu0, 1/sqrt(tau0)) 
prior95 <- qnorm(0.975, mu0, 1/sqrt(tau0))

post05 <- qnorm(0.025, mu1, 1/sqrt(tau1)) 
post95 <- qnorm(0.975, mu1, 1/sqrt(tau1))

@@

Let's illustrate this with an informative prior with $\mu_0$ = `r mu0` and $\tau_0$ = `r tau0`.

```{r}
g_grid <- g(x_grid, mu0, tau0)
outside <- x_grid < prior05 | x_grid > prior95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=g_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

@@

Let's see what happens with this informative prior if you observe `r n` values of x with a mean of `r xbar`. The likelihood is

```{r}
f_grid <- f(xbar, x_grid, tauF)^n
x_color <- "white"
data.frame(x=x_grid, y=f_grid) %>%
  ggplot(aes(x, y)) + 
    geom_line() + xlab(" ") + ylab(" ")
```

@@

```{r}
h_grid <- g(x_grid, mu1, tau1)
outside <- x_grid < post05 | x_grid > post95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=h_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x_grid, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r mu0` and it excludes `r xbar`, the mean of the data.

@@

Define your prior distribution as a normal distributions with prior mean $\mu0$ and prior precision

$\tau_F+\lambda (\tau_0-\tau_F)$

where $\lambda$ ranges between 0 and 1 and $\tau_F$ is less than $\tau_0$. When $\lambda=0$, the prior precision reduces to $\tau_F$ which is equivalent to a prior sample size of $n_0=1$. When $lambda=1$, the prior precision is $\tau_0$ which represents the full strength of the informative priori distribution.

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

gstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF)+tauF)}
density <- outer(mu_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

This is the joint prior distribution of $\mu_0$ and $\lambda$.

It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\mu$ is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r}
density <- outer(mu_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The likelihood function is constant across all values of $\tau$, so its surface looks like the following.

```{r}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
likelihood <- outer(mu_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
density <- outer(mu_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  mu_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(50, 75, 90)/100,
  xlab="mu",
  ylab="lambda")
```

You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r mu0`) is largely ignored in favor of the mean of the data (`r xbar`).

>>>>

Consider a Bayesian analysis of an outcome variable that has a normal distribution.

For this derivation, it is easier to write the density in terms of the precision

$f(X; \mu, \theta) = \sqrt{\frac{\theta}{2 \pi}}e^{-\frac{1}{2}\theta (X-\mu)^2}$

@@

$\theta=\frac{1}{\sigma*2}$

Assume that the mean of the normal distribution, $\mu$, is known and equal to $\mu_F$. An earlier case with precision known and the mean  unknown appears on a different web page. A more general case, with both the mean and the precision of the normal distribution unknown is also described on a separate web page. For simplicity, we will drop any multiplicative constants not involving X or $\theta$

$f(X; \theta) \propto \sqrt{\theta}e^{-\frac{1}{2}\theta(X-\mu_F)^2}$

@@

$g(\theta; \alpha_0, \beta_0) = \frac{\beta_0^{\alpha_0}}{\Gamma(\alpha_0)} \theta^{\alpha_0-1} e^{-\beta_0 \theta}$

Set the prior parameters to

$\alpha_0=n_0$

$\beta_0 = n_0 \theta_0$

and remove any multiplicative constants with respect to $\theta$ to get

$g(\theta; n_0, \theta_0) \propto \theta^{n_0-1} e^{-n_0 \theta_0 \theta}$

In this parameterization, $n_0$ is the prior sample size.

@@

$L(\mathbf{X} | \mu) \propto \theta^{n/2}e^{- \frac{1}{2} \theta\sum(X_i-\mu_0)^2}$

@@

$h(\theta | \mathbf{X}) \propto \theta^{n_0-1} e^{-n_0 \theta / \theta0} \theta^{n/2} e^{- \frac{1}{2} \theta\sum(X_i-\mu_0)^2}$

which can be rewritten as

$h(\theta | \mathbf{X}) \propto \theta^{n/2 + n_0-1} e^{-(n_0 \theta0 + \frac{1}{2}\sum(X_i-\mu_0)^2)}$

which is a gamma distribution with posterior parameters

$\alpha_1 = n/2+n_0$ and

$\beta_1 = n_0 theta_0 + \frac{1}{2}\sum(X_i-\mu_0)^2)$

This can also be shown to relate to a t-distribution

$\frac{\Gamma \left(\frac{\nu+1}{2} \right)} {\sqrt{\nu\pi}\,\Gamma \left(\frac{\nu}{2} \right)} \left(1+\frac{x^2}{\nu} \right)^{-\frac{\nu+1}{2}}$

@@

n0 <- 4
theta0 <- 0.5

alpha0 <- n0
beta0 <- n0*theta0

muF <- 2
n <- 10

x <- muF+ round(rnorm(n), 1)/theta0

ss <- sum((x-muF)^2)

alpha1 <- alpha0 + n
beta1 <- beta0 + 0.5*ss

@@

f <- function(x, mu, theta) {
  sqrt(theta/2*pi)*exp(-0.5*theta*(x-mu)^2)
}

g <- function(x, alpha, beta) {
  beta^alpha/gamma(alpha)*x^(alpha-1)*exp(-beta*x)
}

prior05 <- qgamma(0.025, alpha0, beta0) 
prior95 <- qgamma(0.975, alpha0, beta0)

post05 <- qgamma(0.025, alpha1, beta1)  
post95 <- qgamma(0.975, alpha1, beta1)

@@

Let's illustrate this with an informative prior with $n_0$ = `r n0` and $\theta_0$ = `r theta0`.

```{r}
g_grid <- g(x_grid, alpha0, beta0)
outside <- x_grid < prior05 | x_grid > prior95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=g_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

@@

Let's see what happens with this informative prior if you observe `r n` values of x: `r paste0(x, collapse=", ")`. The likelihood is

```{r}
f_grid <- rep(NA, length(x_grid))
for (i in 1:length(x_grid)) {
  f_grid[i] <- prod(f(x, muF, x_grid[i]))
}
x_color <- "white"
data.frame(x=x_grid, y=f_grid) %>%
  ggplot(aes(x, y)) + 
    geom_line() + xlab(" ") + ylab(" ")
```

@@

```{r}
h_grid <- g(x_grid, alpha1, beta1)
outside <- x_grid < post05 | x_grid > post95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=h_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x_grid, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r theta0` and it excludes `r ss/(n-1)`, the mean of the data.

@@

Define your prior distribution as a gamma distributions with prior

mean $\theta_0$ and prior precision

$\alpha = 1 + \lambda(n0-1)$

$\beta = \alpha \theta_0 $

where $\lambda$ ranges between 0 and 1. When $\lambda=0$, the prior precision reduces to $alpha=1$ which is equivalent to a very diffuse prior. When $lambda=1$, the prior precision is $\alpha=\alpha_0$ or the full strength prior.

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

gstar <- function(theta, lambda) {g(theta, 1+lambda*(alpha0-1), (1+lambda*(alpha0-1))*beta0)}
density <- outer(theta_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

This is the joint prior distribution of $\theta$ and $\lambda$.

It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\theta$ is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r, eval=FALSE}
density <- outer(theta_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The likelihood function is constant across all values of $\theta$, so its surface looks like the following.

```{r, eval=FALSE}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

fstar <- function(mu, lambda) {f(xbar, mu, theta0)^n}
likelihood <- outer(mu_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r, eval=FALSE}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)

hstar <- function(mu, lambda) {g(mu, mu0, lambda*(theta0-thetaF) + thetaF)*f(xbar, mu, thetaF)^n}
density <- outer(mu_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r, eval=FALSE}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  mu_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(50, 75, 90)/100,
  xlab="theta",
  ylab="lambda")
```

You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r theta0`) is largely ignored in favor of the mean of the data (`r "??"`).
